---
title: "Biaxial Gating of a Single Sample (Fig. S3)"
author: "Benjamin R Babcock"
output: rmarkdown::html_vignette
description: >
  How are cells classified in a single sample? Single-sample cell classification is a key part of generating a Cell Misclassification Statistic. This vignette describes exactly how cells from a single sample/dataset are categorized as part of a multi-dataset workflow.
vignette: >
  %\VignetteIndexEntry{Biaxial Gating of a Single Sample (Fig. S3)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r packages, include=F}
library(knitr)
library(dplyr)
library(Seurat)
library(ggplot2)

opts_chunk$set(
  fig.pos = "!h", out.extra = "", warning = F, 
  fig.align = "center", fig.height = 7, fig.width = 7
)
```
# Processing the Seurat Object:
## Data filtering & normalization
```{r inital_processing, message = FALSE}
library(BatchNorm)

# Import unfiltered Seurat object (included with 'BatchNorm' package)
data(PBMC4A)

# Run "standard" Seurat workflow
PBMC4A <- PBMC4A %>% 
          MitoFilter() %>% 
          NormalizeData(assay = "RNA", verbose = FALSE) %>%
          FindVariableFeatures(selection.method = "vst", nfeatures = 2000, verbose = FALSE) %>%
          ScaleData(verbose = FALSE) %>% 
          RunPCA(npcs = 30, verbose = FALSE)
```
## Selecting the appropriate number of Principal Components for UMAP reduction
```{r PC_Selection, message = FALSE}
# Identify correct numbers of PCs
#PBMC4A.pca.test <- TestPCA(object = PBMC4A)
#PBMC4A.pca.test[, 1:20]
# 16 PCs with a z-score > 1
# Proceed with 16 PCs for dimensional reduction & clustering
# Visualize PCs plotted by standard deviation:
ElbowPlot(PBMC4A)
```
## Selecting the appropriate number of Principal Components for UMAP reduction
```{r Clustering, message = FALSE}
PBMC4A <- PBMC4A %>% 
          RunUMAP(reduction = "pca", dims = 1:16, verbose = FALSE) %>%
          FindNeighbors(reduction = "pca", dims = 1:16, verbose = FALSE) %>%
          FindClusters(resolution = .8, verbose = FALSE)
# Visualize clusters on UMAP
DimPlot(PBMC4A, reduction = "umap", cols = colors.use) + 
        ggtitle("PBMC Clusters")
```

# Biaxial Antigen Gating

## B Cell Classification
```{r B Cell Classification, message = F}
# Produce scatter plot of CD20 vs CD19
d <- as.data.frame(FetchData(object = PBMC4A, 
                   vars = c("adt_CD20", "adt_CD19", "seurat_clusters")))
## Format colnames to match FeatureScatter syntax
colnames(d) <- c("adt_CD20", "adt_CD19", "colors")
FeatureScatter(PBMC4A, "adt_CD20", "adt_CD19",
               cols = colors.use, plot.cor = F) +
     geom_density2d(data = d, color = "black", size = 1) +
     theme_gray() + 
     labs(x = "CD20 (ADT)", 
          y = "CD19 (ADT)")
# clusters 6 & 9 contain B cells


# Subset B cells (CD20 vs CD19) and isolate Naive B cells (IGD+/CD27-)
bcells <- subset(PBMC4A, `adt_CD20` > 25 | `adt_CD19` > 10)
bcells[["IGD"]] <- colSums(bcells@assays$RNA@data[grep("^IGHD", rownames(bcells)), ])

d <- as.data.frame(FetchData(object = bcells, 
                   vars = c("IGD", "adt_CD27", "seurat_clusters")))
## Format colnames to match FeatureScatter syntax
colnames(d) <- c("IGD", "adt_CD27", "colors")
FeatureScatter(bcells, "adt_CD27", "IGD", plot.cor = F,
                    cols = colors.use[c(2, 4, 7, 9, 10, 14, 15)]) +
     geom_density2d(data = d, color = "black", size = 1) +
     labs(x = "CD27 (ADT)", 
          y = "Total IGHD Expression (RNA)") + theme_gray() +
     coord_cartesian(ylim = c(-0.3, 3.5), expand = T)
# cluster 6 contains Naive B cells
```
## NK Cell Classification
```{r}
# Remove B cells (already classified) and continue to classify non-B cell PBMCs
nonclassified <- subset(PBMC4A, idents = c(6, 9), invert = T)
d <- as.data.frame(FetchData(object = nonclassified, 
                   vars = c("adt_CD56", "adt_CD14", "seurat_clusters")))
## Format colnames to match FeatureScatter syntax
colnames(d) <- c("adt_CD56", "adt_CD14", "colors")
FeatureScatter(nonclassified, "adt_CD14", "adt_CD56", plot.cor = F, 
                    cols = colors.use[c(1:6, 8:9, 11:17)], pt.size = 3) +
     geom_density2d(data = d, color = "black", size = 1) +
     labs(x = "CD14 (ADT)", 
          y = "CD56 (ADT)") + theme_gray()
# clusters 3, 4, 5 & 12 appear to contain NK cells
# Cluster 12 appears to contain mostly CD56Hi NK cells
# We can further classify NK cells and confirm using CD3 ADT & NKG7 transcript

nkcells <- subset(PBMC4A, idents = c(3, 4, 5, 12))
d <- as.data.frame(FetchData(object = nkcells, 
                   vars = c("NKG7", "adt_CD3", "seurat_clusters")))
## Format colnames to match FeatureScatter syntax
colnames(d) <- c("NKG7", "adt_CD3", "colors")
FeatureScatter(nkcells, "adt_CD3", "NKG7", 
                    cols = colors.use[c(4, 5, 6, 14)], pt.size = 3, plot.cor = F) +
     geom_density2d(data = d, color = "black", size = 1) +
     labs(x = "CD3 (ADT)", 
          y = "NKG7 (RNA)") + theme_gray()
# cluster 12 = CD56Hi NK cells, Clusters 3 & 4 = NK Cells, Cluster 5 = NK-T Cells
```
## Monocyte Classification
```{r}
nonclassified <- subset(PBMC4A, idents = c(6, 9, 3, 4, 12, 5), invert = T)

d <- as.data.frame(FetchData(object = nonclassified, 
                   vars = c("adt_CD16", "adt_CD14", "seurat_clusters")))
colnames(d) <- c("adt_CD16", "adt_CD14", "colors")
FeatureScatter(nonclassified, "adt_CD14", "adt_CD16", 
                    cols = colors.use[c(1:3, 8:9, 11:12, 14:17)], pt.size = 3, plot.cor = F) +
     geom_density2d(data = d, color = "black", size = 1) +
     labs(x = "CD14 (ADT)", 
          y = "CD16 (ADT)") + theme_gray()
# Cluster 8 = classical monocytes, 13 = non-classical monocytes
```
## HSPC Classification
```{r}
nonclassified <- subset(PBMC4A, idents = c(6, 9, 3, 4, 12, 5, 8, 13), invert = T)

# Create a variable comparing only cluster 16 to all other unclassified clusters
nonclassified[["HSPC_gate"]] <- rep("Other Clusters", nrow(nonclassified[[]]))
HSPCs <- WhichCells(nonclassified, idents = 16)
nonclassified@meta.data[HSPCs, "HSPC_gate"] <- "16"
nonclassified@meta.data[, "HSPC_gate"] <- factor(nonclassified@meta.data[, "HSPC_gate"],
                                                  levels = c("Other Clusters", "16"))
VlnPlot(nonclassified, "CD34", cols = colors.use[c(1:3, 8, 11:12, 15:17)],
        group.by = "HSPC_gate")
# Cluster 16 = HSPCs
```
## T Cell Classification
```{r}
nonclassified <- subset(PBMC4A, idents = c(6, 9, 3, 4, 12, 8, 5, 13, 16), invert = T)

d <- as.data.frame(FetchData(object = nonclassified, 
                   vars = c("adt_CD8A", "adt_CD4", "seurat_clusters")))
colnames(d) <- c("adt_CD8A", "adt_CD4", "colors")
FeatureScatter(nonclassified, "adt_CD8A", "adt_CD4", 
                    cols = colors.use[c(1:3, 8, 11:12, 15:16)], pt.size = 3, plot.cor = F) +
     geom_density2d(data = d, color = "black", size = 1) +
     labs(x = "CD8A (ADT)", 
          y = "CD4 (ADT)") + theme_gray()
# 0, 1, 2, 15 = CD4+ T-cell; 7, 11 = CD8+ T-cell

nonclassified <- subset(PBMC4A, idents = c(6, 9, 3, 4, 12, 8, 5, 13, 16, 7, 11), invert = T)
# Create a variable comparing only cluster 10 to all other unclassified clusters (other CD4+ T Cells)
nonclassified[["Treg_gate"]] <- rep("Other CD4+", nrow(nonclassified[[]]))
Tregs <- WhichCells(nonclassified, idents = 10)
nonclassified@meta.data[Tregs, "Treg_gate"] <- "10"
nonclassified@meta.data[, "Treg_gate"] <- factor(nonclassified@meta.data[, "Treg_gate"],
                                                  levels = c("Other CD4+", "10"))
VlnPlot(nonclassified, "FOXP3", cols = colors.use[c(11, 15)],
        group.by = "Treg_gate")

# Cluster 10 = Treg cells
```
## Plasmacytoid Dendritic Cell Classification
```{r}
nonclassified <- subset(PBMC4A, idents = c(6, 9, 3, 4, 12, 8, 5, 13, 16, 7, 11), invert = T)
nonclassified[["pDC_gate"]] <- rep("Other CD4+", nrow(nonclassified[[]]))
PDCs <- WhichCells(nonclassified, idents = 14)
nonclassified@meta.data[PDCs, "pDC_gate"] <- "14"
nonclassified@meta.data[, "pDC_gate"] <- factor(nonclassified@meta.data[, "pDC_gate"],
                                                  levels = c("Other CD4+", "14"))

VlnPlot(nonclassified, 'LILRA4', cols = colors.use[c(11, 15)],
        group.by = "pDC_gate")
# Cluster 14 = pDCs
```
# Assigning Cell Classification Labels
```{r}
# Assign cell classification labels to clusters, as follows:

 # 6 = B Naive
 # 9 = B memory
 # 3, 4 = NK
 # 5 = NK_T
 # 12 = CD56Hi NK
 # 8 = classical monocytes 
 # 13 = non-classical monocytes
 # 16 = HSPCs
 # 0, 1, 2, 15 = CD4+ T-cell
 # 7, 11 = CD8+ T-cell
 # 10 = TRegs
 # 14 = DCs
Idents(PBMC4A) <- PBMC4A[["seurat_clusters"]]
Idents(PBMC4A) <- plyr::mapvalues(Idents(PBMC4A), from = c(0, 1, 2, 3, 4, 5,
                                                   6, 7, 8, 9, 10,
                                                   11, 12, 13, 14, 15,
                                                   16), 
                to = c('T_CD4', 'T_CD4', 'T_CD4', 'NK', 'NK', 'NK_T',
                       'B_Naive', 'T_CD8', 'Monocyte_Classical', 'B_Memory', 'TReg',
                       'T_CD8', 'NK_CD56Hi', 'Monocyte_NonClassical', 'Dendritic_Cells', 'T_CD4',
                       'HSPCs'))
# Order cell classifications
Idents(PBMC4A) <- factor(Idents(PBMC4A),
                                 levels = c("B_Naive", "B_Memory", "T_CD4", "TReg", 
                                            "T_CD8", "NK_T", "NK", "NK_CD56Hi",
                                            "Monocyte_Classical", "Monocyte_NonClassical",
                                            "Dendritic_Cells", "HSPCs"))
PBMC4A[["Cell_Type"]] <- Idents(PBMC4A)
UMAPPlot(PBMC4A, cols = colors.use, pt.size = 2, label = F)
```

